Clustering NBA players utilizing K-Means, Gaussian Mixture Models, Principal Component Analysis and Graphical Networks


First we will cluster types of shooters

```{r}
# 1. Load Libraries
library(dplyr)
library(tidyr)
library(factoextra)

# 2. Load the Local CSVs
# Make sure your working directory is set to where you saved the files
df_adv <- read.csv("ad.csv", stringsAsFactors = FALSE)
df_per100 <- read.csv("pos.csv", stringsAsFactors = FALSE)
df_shoot <- read.csv("shot.csv", stringsAsFactors = FALSE)

# 3. Clean Duplicates (Keep 'TOT' row for traded players)
clean_duplicates <- function(df) {
  df %>%
    group_by(Player) %>%
    # If a player has multiple rows, keep the one where Team is 'TOT' (Total)
    # If they only have one row, keep it.
    filter(n() == 1 | Team == "2TM") %>%
    ungroup()
}

df_adv <- clean_duplicates(df_adv)
df_per100 <- clean_duplicates(df_per100)
df_shoot <- clean_duplicates(df_shoot)

# 4. Merge the Tables Together
# We join by Player, Pos, and Age to ensure perfectly matched rows
merged_data <- df_adv %>%
  inner_join(df_per100, by = c("Player", "Pos", "Age", "Team", "G")) %>%
  inner_join(df_shoot, by = c("Player", "Pos", "Age", "Team", "G"))

# 5. Select Your Exact Features & Filter Minutes
# Check your CSV headers. Bball-Ref usually names columns like "X0.3" for 0-3 feet when loaded into R.
# The names below are the standard outputs from their CSV export.
shot_data <- merged_data %>%
  # Filter by MP.x (Minutes Played from the advanced table) to remove tiny samples
  filter(as.numeric(MP.x) >= 500) %>%
  select(
    Player,
    
    # --- ROLE & VOLUME ---
    USG. = USG.,          # Usage Percentage (Primary vs Role Player)
    FGA_per100 = FGA,     # Total Shot Volume 
    AST_per100 = AST,     # Playmaking / Ball Dominance
    
    # --- SHOT PROFILE (WHERE) ---
    Dist = Dist. ,          # Average Shot Distance
    X3PAr = X3PAr,        # 3-Point Attempt Rate
    pct_FGA_0_3 = X0.3,   # % of FGA from 0-3 ft (Paint Scorers)
    pct_FGA_3_10 = X3.10, # % of FGA from 3-10 ft
    pct_FGA_10_16 = X10.16, #  Long Mid-Range
    pct_FGA_3P = X3PAr,     # % of FGA from 3PT (Floor Spacers)
    
    # --- EFFICIENCY / SKILL  ---
    fg_pct_0_3 = X0.3.1,   # FG% at the rim
    fg_pct_3_10 = X3.10.1, # FG% from 3-10 ft
    fg_pct_mid = X10.16.1, # FG% from 10-16 ft
    fg_pct_long_mid = X16.3P.1, #FG% 16ft-3pt
    fg_pct_3P = X3P.1,     #  3P%
    
    # --- SPECIALTY ZONES ---
    pct_3PA_Corner = X.3PA,  # Corner 3s
    pct_FGA_Dunks = X.FGA,

    
    # Shot Creation (Assisted %)
    pct_Ast_2P = X2P.2,   # % of 2P Field Goals Assisted
    pct_Ast_3P = X3P.2    # % of 3P Field Goals Assisted
  ) %>%
  # Convert all metric columns to numeric (in case they loaded as characters)
  mutate(across(-Player, as.numeric)) %>%
  #drop_na()
  mutate(across(everything(), ~replace_na(.x, 0)))

# 6. Scale Data (Dropping the Player name column)
scaled_data <- scale(shot_data[, -1])

# 7. Find Optimal K (Plot the elbow)
fviz_nbclust(scaled_data, kmeans, method = "wss")+ 
  labs(title = "Elbow Method for Shooter Archetypes")

```
Not using a library and doing the for loop explicitly

```{r}
# 7. Find Optimal K (Manual SSE Loop)
set.seed(222) 
MAX_K <- 10 # Capped at 10 to prevent micro-clusters
sse <- c() 

# Make sure 'scaled_offense' is the matrix you are feeding into X
X <- scaled_data

for (k in 1:MAX_K) {
  algo_k <- kmeans(X, centers=k, nstart=22, iter.max=20) 
  sse <- c(sse, algo_k$tot.withinss) 
}

# Plot the elbow
plot(1:MAX_K, sse, type="b", pch=19, col="blue",
     xlab="Number of clusters K", 
     ylab="Total within-clusters sum of squares (SSE)")
```

This now takes the 1st derivative

```{r}
tibble(k = 1:MAX_K, SSE_difference = sse-lead(sse)) %>%
  dplyr::filter(k<MAX_K) %>%
  ggplot(aes(x=k, y=SSE_difference)) + 
  geom_point(color="#F84C1E") + geom_line(color="#232D4B") + # set color of point and lines
  labs(x = "K", y = "SSE Rolling Difference", title = "A Clearer Picture") + # set axis/plot titles
  scale_x_continuous(breaks=seq(1, MAX_K, 1)) + # define x-axis
  theme(panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank()) # manually alter theme
```

Now we will take the 2nd derivative

```{r}
tibble(k = 1:MAX_K, SSE_difference = sse-2*lead(sse)+lead(sse, 2)) %>%
  dplyr::filter(k<MAX_K-1) %>%
  ggplot(aes(x=k, y=SSE_difference)) + 
  geom_point(color="#F84C1E") + geom_line(color="#232D4B") + 
  labs(x = "K", y = "SSE Rolling 2-Unit Difference", title = "An Even Clearer Picture") + 
  scale_x_continuous(breaks=seq(1, MAX_K, 1)) + 
  theme(panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank())
```
I will go with k=4 clusters

```{r}
# 1. Run Final K-Means with K = 5
set.seed(222) # Keeping your seed for reproducibility [cite: 1]
final_shooter_kmeans <- kmeans(scaled_data, centers = 5, nstart = 25)

# 2. Attach the cluster labels back to your original unscaled data
shot_data$Shooter_Cluster <- as.factor(final_shooter_kmeans$cluster)

# 3. Calculate the average stats for each cluster to see what they represent
cluster_profiles <- shot_data %>%
  group_by(Shooter_Cluster) %>%
  summarise(
    Count = n(),
    Avg_USG = round(mean(USG.), 1),
    Avg_AST_per100 = round(mean(AST_per100), 1),
    Avg_Dist = round(mean(Dist), 1),
    Avg_3PAr = round(mean(X3PAr), 2),
    Avg_0_3ft = round(mean(pct_FGA_0_3), 3),
    Avg_Dunks = round(mean(pct_FGA_Dunks), 3),
    Avg_Ast_2P = round(mean(pct_Ast_2P), 3),
    Avg_Ast_3P = round(mean(pct_Ast_3P), 3)
    AVG_
  )

# Print the profiles to the console
print(cluster_profiles)

# 4. Export the mapping for Python / PyTorch
# We only need the Player name and their Cluster ID
shooter_export <- shot_data %>%
  select(Player, Shooter_Cluster)

write.csv(shooter_export, "shooter_archetypes_15_16.csv", row.names = FALSE)
```


```{r}
# 1. Crear el vector de nombres de columnas (asegúrate de que coincida con tus datos)
features_names <- c("USG.", "FGA_per100", "AST_per100", "Dist", "X3PAr", 
                    "pct_FGA_0_3", "pct_FGA_3_10", "pct_FGA_10_16", "pct_FGA_3P", 
                    "pct_3PA_Corner", "pct_FGA_Dunks", "pct_Ast_2P", "pct_Ast_3P")

# 2. Asignar nombres a las columnas de los centros y añadir la columna Cluster
km_centers <- as.data.frame(final_shooter_kmeans$centers)
colnames(km_centers) <- features_names # Asigna los nombres al DF
km_centers$Cluster <- c('Cluster 1', 'Cluster 2', 'Cluster 3', 'Cluster 4', 'Cluster 5')

# 3. Pivotar a formato largo para que ggplot reconozca 'feature' y 'z_val'
km_centers_long <- km_centers %>% 
  pivot_longer(cols = -Cluster, names_to = "feature", values_to = "z_val")

# 4. Tu código de ggplot (usando km_centers_long)
km_centers_long %>% 
  ggplot(aes(x = feature, y = z_val, color = Cluster)) + 
  geom_point() + 
  scale_color_brewer(palette = "Paired") + 
  gghighlight(use_direct_label = FALSE) + 
  facet_wrap(~ Cluster, ncol = 3) + 
  labs(x = "Predictor", y = "Cluster Center (Z-score)", 
       title = "Visualizing K-Means Cluster Makeups") + 
  theme_minimal() + # Sugerido para limpieza visual
  theme(legend.position = "none", 
        strip.text = element_text(face = 'bold'),
        axis.text.x = element_text(angle = 90, size = 8, vjust = 0.5, hjust = 1), 
        panel.grid.minor = element_blank())
```


```{r}
pca <- prcomp(scaled_data) # perform Principle Component Analysis 
pca_summary <- summary(pca) # summary of PCA model

# plot % of variance between players explained by each subsequent PC 
tibble(imp = pca_summary$importance[2,], n = 1:length(imp)) %>% # get importance scores for PCA summary
  ggplot(aes(x=n, y=imp)) + 
  labs(x = 'Principle Component #', y = '% of Variance Explained by Component',
       title = 'Less Information is Gained by Each Subsequent PC') +
  geom_point(color="#F84C1E") + geom_line(color="#232D4B") + 
   scale_x_continuous(breaks=seq(1, 20, 1)) + # set x-axis
  scale_y_continuous(labels=scales::percent) + # change y-axis from proportion to percentage
  theme(panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank())
```
```{r}
pc2 <- as.data.frame(pca$x[,1:2]) # extract first 3 PCs
pc2$Cluster <- as.factor(final_shooter_kmeans$cluster) # add player clusters 
cluster1_var <- round(pca_summary$importance[2,1], 4) * 100 # get variance explained by cluster 1
cluster2_var <- round(pca_summary$importance[2,2], 4) * 100 # get variance explained by cluster 2

# how different are the clusters when scaled down to two dimensions? 
pc2 %>% 
  ggplot(aes(x=PC1, y=PC2, color=Cluster, shape=Cluster)) + 
  geom_point(alpha=0.3) + 
  scale_color_futurama() + # fun color scale
  geom_rug() + # great way to visualize points on a single axis
  stat_ellipse(level=(2/3)) + # set ellipse value to one standard deviation
  scale_shape_manual(values=seq(0,15)) + 
  labs(x = paste0('PC1 (Accounts for ', cluster1_var, '% of Variance)'), # define cluster 1 % of variance
       y = paste0('PC2 (Accounts for ', cluster2_var, '% of Variance)'), # define cluster 2 % of variance
       title = 'Visualizing K-Means Cluster Differences in 2D') 
```
Gaussian mixture model


```{r}
# run GMM on the scaled data from above
set.seed(222)
mixture_model <-  Mclust(pca$x[,1:3])
mix_model_summary <- summary(mixture_model) # obtain model summary
mix_model_summary
```
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# 1. Get the MAP class labels from your PCA-trained GMM
gmm_labels <- mixture_model$classification

# 2. Bind those labels to your ORIGINAL scaled data (so we have basketball stats, not PCs)
gmm_profiles <- as.data.frame(scaled_data) %>%
  mutate(Cluster = paste("Cluster", gmm_labels))

# 3. Calculate the true centers (means) for each feature by Cluster
mbc_centers <- gmm_profiles %>%
  group_by(Cluster) %>%
  summarise(across(everything(), mean))

# 4. Pivot for plotting (exactly as you wrote it!)
mbc_centers_long <- mbc_centers %>%
  pivot_longer(!Cluster, names_to = 'feature', values_to = 'z_val')

# 5. Plot the Z-Score Profiles [This creates a beautiful line chart of your archetypes]
ggplot(mbc_centers_long, aes(x = feature, y = z_val, color = Cluster, group = Cluster)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") + # 0 is the NBA average
  theme_minimal() +
  labs(title = "GMM Cluster Profiles (Z-Scores)",
       x = "Offensive Feature",
       y = "Standard Deviations from NBA Average") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# changing the .class assignment in this chunk of code allows you to explore the players with the highest probability of belonging to that cluster
mix_model_augment <- augment(mixture_model, shot_data)
mix_model_augment %>%
  dplyr::select(Player, .class, .uncertainty) %>% # select columns describing the player / cluster
  dplyr::filter(.class==1) %>% # filter to an individual cluster
  arrange(-.uncertainty) %>% # sort by probability
  head(5)
```


```{r}
pc2 <- as.data.frame(pca$x[,1:2]) # extract first two PCs 
pc2$Cluster <- as.factor(mix_model_augment$.class) # add player clusters 


pc2 %>% 
  ggplot(aes(x=PC1, y=PC2, color=Cluster, shape=Cluster)) + 
  geom_point(alpha=0.5) +
  scale_color_manual(values = pal_npg()(7)) + 
  geom_rug() + theme_minimal() + 
  stat_ellipse(level=(2/3)) + # set ellipse value to one standard deviation
  scale_shape_manual(values=seq(0,7)) +
  labs(x = paste0('PC1 (Accounts for ', cluster1_var, '% of Variance)'),
       y = paste0('PC2 (Accounts for ', cluster2_var, '% of Variance)'),
       title = 'Visualizing Model Based Cluster Differences in 2D') 
```

```{r}

mma <- mix_model_augment %>%
  dplyr::select( nPlayer, .class, .uncertainty) %>%
  cbind(dplyr::select(mix_model_preds, -classification)) %>% 
```


```{r}
library(dplyr)
library(broom)

mma <- mix_model_augment %>%
  dplyr::select(Player, .class, .uncertainty) 
  #cbind(dplyr::select(mix_model_preds, -classification))  # get soft classification probabilities for each cluster

# 1. Create your nodes dataframe (using tibble instead of the deprecated data_frame)
nodes <- tibble(
  id = mma$Player, 
  group = mma$.class
)

# 2. Extract ONLY the cluster probability columns to a matrix
# (Adjust the select logic if your probability columns are named differently)
prob_matrix <- mma %>% 
  select(-Player, -.class, -.uncertainty) %>% 
  as.matrix()

# Assign player names to the rows so the distance matrix knows who is who
rownames(prob_matrix) <- mma$Player

# 3. Calculate all pairwise Euclidean distances instantly
dist_mat <- dist(prob_matrix, method = "euclidean")

# 4. Convert the distance matrix into an edge list!
# broom::tidy() automatically creates a clean 'item1', 'item2', 'distance' dataframe
edges <- tidy(dist_mat) %>%
  rename(from = item1, to = item2, weight = distance)

# (Optional but recommended): Invert the weight for network graphing.
# In networks, "weight" usually means connection strength (closer = higher weight).
# Right now, a distance of 0 means they are identical. 
edges <- edges %>%
  mutate(similarity_weight = 1 - weight)
```




Mnimum spanning tree

```{r}
# Install these if you haven't already: install.packages(c("igraph", "ggraph"))
library(igraph)
library(ggraph)

# 1. Clean up the edges dataframe for igraph
# igraph expects the first two columns to be 'from' and 'to'.
# We want to use the original Euclidean distance as the 'weight' because 
# a Minimum Spanning Tree looks for the *smallest* total weight (shortest distance = most similar).
edges_clean <- edges %>%
  dplyr::select(from, to, weight) %>%
dplyr::filter(is.finite(weight))

# 2. Convert your dataframes into an igraph network object
full_graph <- graph_from_data_frame(d = edges_clean, vertices = nodes, directed = FALSE)

# 3. Calculate the Minimum Spanning Tree! 
# This instantly slices your 60,000+ edges down to exactly N-1 edges.
mst_graph <- mst(full_graph)

# 4. Plot the stunning result using ggraph
set.seed(222) # For a reproducible layout
ggraph(mst_graph, layout = 'tree') + # 'kk' (Kamada-Kawai) is a great physics-based layout for MSTs
  geom_edge_link(alpha = 0.3, color = "darkgray") +
  geom_node_point(aes(color = as.factor(group)), size = 3) +
  # Optional: Add text labels for the most extreme players on the edges of the tree
  # geom_node_text(aes(label = name), repel = TRUE, size = 3, check_overlap = TRUE) + 
  theme_void() +
  labs(title = "Minimum Spanning Tree of NBA Shooter Archetypes",
       subtitle = "Connected by shortest multi-dimensional distance",
       color = "GMM Cluster") +
  theme(legend.position = "bottom")
```




Lets now compare the Gaussian model with kmeans
```{r}
shot_data$KMeans_Cluster <- as.factor(final_shooter_kmeans$cluster)

# 2. Attach your new GMM clusters from the PCA
shot_data$GMM_Cluster <- as.factor(mixture_model$classification)

# 3. Create the cross-tabulation table to compare them
comparison_table <- table(KMeans = shot_data$KMeans_Cluster, 
                          GMM = shot_data$GMM_Cluster)

# Print the table
print(comparison_table)
```










