Clustering NBA players utilizing K-Means, Gaussian Mixture Models, Principal Component Analysis 


First we will cluster types of shooters

```{r}
# 1. Load Libraries
library(dplyr)
library(tidyr)
library(factoextra)
library(ggplot2)

# 2. Load the Local CSVs
# Make sure your working directory is set to where you saved the files
df_adv <- read.csv("data/ad.csv", stringsAsFactors = FALSE)
df_per100 <- read.csv("data/pos.csv", stringsAsFactors = FALSE)
df_shoot <- read.csv("data/shot.csv", stringsAsFactors = FALSE)

# 3. Clean Duplicates (Keep 'TOT' row for traded players)
clean_duplicates <- function(df) {
  df %>%
    group_by(Player) %>%
    # If a player has multiple rows, keep the one where Team is 'TOT' (Total)
    # If they only have one row, keep it.
    filter(n() == 1 | Team == "2TM") %>%
    ungroup()
}

df_adv <- clean_duplicates(df_adv)
df_per100 <- clean_duplicates(df_per100)
df_shoot <- clean_duplicates(df_shoot)

# 4. Merge the Tables Together
# We join by Player, Pos, and Age to ensure perfectly matched rows
merged_data <- df_adv %>%
  inner_join(df_per100, by = c("Player", "Pos", "Age", "Team", "G")) %>%
  inner_join(df_shoot, by = c("Player", "Pos", "Age", "Team", "G"))

# 5. Select Your Exact Features & Filter Minutes
# Check your CSV headers. Bball-Ref usually names columns like "X0.3" for 0-3 feet when loaded into R.
# The names below are the standard outputs from their CSV export.
shot_data <- merged_data %>%
  # Filter by MP.x (Minutes Played from the advanced table) to remove tiny samples
  filter(as.numeric(MP.x) >= 500) %>%
  select(
    Player,
    
    # --- ROLE & VOLUME ---
    USG. = USG.,          # Usage Percentage (Primary vs Role Player)
    #FGA_per100 = FGA,     # Total Shot Volume 
   # AST_per100 = AST,     # Playmaking / Ball Dominance
    
    # --- SHOT PROFILE (WHERE) ---
   # Dist = Dist. ,          # Average Shot Distance
    X3PAr = X3PAr,        # 3-Point Attempt Rate
    #pct_FGA_0_3 = X0.3,   # % of FGA from 0-3 ft (Paint Scorers)
    #pct_FGA_3_10 = X3.10, # % of FGA from 3-10 ft
    #pct_FGA_10_16 = X10.16, #  Long Mid-Range

    # --- EFFICIENCY / SKILL  ---
    fg_pct_0_3 = X0.3.1,   # FG% at the rim
    fg_pct_3_10 = X3.10.1, # FG% from 3-10 ft
    fg_pct_mid = X10.16.1, # FG% from 10-16 ft
    #fg_pct_long_mid = X16.3P.1, #FG% 16ft-3pt
    fg_pct_3P = X3P.1,     #  3P%
    
    # --- SPECIALTY ZONES ---
    #pct_3PA_Corner = X.3PA,  # Corner 3s
    #pct_FGA_Dunks = X.FGA,

    
    # Shot Creation (Assisted %)
    FTr = FTr,
    pct_Ast_2P = X2P.2,   # % of 2P Field Goals Assisted
    pct_Ast_3P = X3P.2    # % of 3P Field Goals Assisted
  ) %>%
  # Convert all metric columns to numeric (in case they loaded as characters)
  mutate(across(-Player, as.numeric)) 
  #drop_na()
 # mutate(across(everything(), ~replace_na(.x, 0)))
  shot_data <- shot_data %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), mean(., na.rm = TRUE), .)))
  
  shot_data <- shot_data %>%
  mutate(
    # Pull the extreme foul-drawers closer to the mean
    FTr = sqrt(FTr)
  )

# 6. Scale Data (Dropping the Player name column)
scaled_data <- scale(shot_data[, -1])


```



Do some PCA analysis
```{r}
pca <- prcomp(scaled_data) # perform Principle Component Analysis 
pca_summary <- summary(pca) # summary of PCA model

# plot % of variance between players explained by each subsequent PC 
tibble(imp = pca_summary$importance[2,], n = 1:length(imp)) %>% # get importance scores for PCA summary
  ggplot(aes(x=n, y=imp)) + 
  labs(x = 'Principle Component #', y = '% of Variance Explained by Component',
       title = 'Less Information is Gained by Each Subsequent PC') +
  geom_point(color="#F84C1E") + geom_line(color="#232D4B") + 
   scale_x_continuous(breaks=seq(1, 20, 1)) + # set x-axis
  scale_y_continuous(labels=scales::percent) + # change y-axis from proportion to percentage
  theme(panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank())
```

What does eac PC mean
```{r}

# 1. Extract the rotation matrix (the loadings) for the first 3 PCs
pca_loadings <- as.data.frame(pca$rotation[, 1:6])
pca_loadings$Feature <- rownames(pca_loadings)

write.csv(pca_loadings, "pca_loadings.csv", row.names = FALSE)


# 2. Pivot the data so we can plot it easily
loadings_long <- pca_loadings %>%
  pivot_longer(cols = c(PC1, PC2, PC3,PC4,PC5,PC6), names_to = "PC", values_to = "Loading")

# 3. Plot the specific "recipes" for PC1, PC2, and PC3
ggplot(loadings_long, aes(x = reorder(Feature, Loading), y = Loading, fill = Loading > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~PC) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "darkred")) +
  theme_minimal() +
  labs(title = "PCA Loadings: The DNA of the Principal Components",
       x = "Offensive Stat",
       y = "Weight (Loading)")
```







Running k-means 
```{r}
# 7. Find Optimal K (Plot the elbow)
fviz_nbclust(scaled_data, kmeans, method = "wss")+ 
  labs(title = "Elbow Method for Shooter Archetypes")
```


Not using a library and doing the for loop explicitly we do this to be able to calculate the 1st and 2nd derivative

```{r}
# 7. Find Optimal K (Manual SSE Loop)
set.seed(222) 
MAX_K <- 10 # Capped at 10 to prevent micro-clusters
sse <- c() 

# Make sure 'scaled_offense' is the matrix you are feeding into X
X <- scaled_data

for (k in 1:MAX_K) {
  algo_k <- kmeans(X, centers=k, nstart=22, iter.max=300) 
  sse <- c(sse, algo_k$tot.withinss) 
}

# Plot the elbow
plot(1:MAX_K, sse, type="b", pch=19, col="blue",
     xlab="Number of clusters K", 
     ylab="Total within-clusters sum of squares (SSE)")
```

This now takes the 1st derivative

```{r}
tibble(k = 1:MAX_K, SSE_difference = sse-lead(sse)) %>%
  dplyr::filter(k<MAX_K) %>%
  ggplot(aes(x=k, y=SSE_difference)) + 
  geom_point(color="#F84C1E") + geom_line(color="#232D4B") + # set color of point and lines
  labs(x = "K", y = "SSE Rolling Difference", title = "A Clearer Picture") + # set axis/plot titles
  scale_x_continuous(breaks=seq(1, MAX_K, 1)) + # define x-axis
  theme(panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank()) # manually alter theme
```

Now we will take the 2nd derivative

```{r}
tibble(k = 1:MAX_K, SSE_difference = sse-2*lead(sse)+lead(sse, 2)) %>%
  dplyr::filter(k<MAX_K-1) %>%
  ggplot(aes(x=k, y=SSE_difference)) + 
  geom_point(color="#F84C1E") + geom_line(color="#232D4B") + 
  labs(x = "K", y = "SSE Rolling 2-Unit Difference", title = "An Even Clearer Picture") + 
  scale_x_continuous(breaks=seq(1, MAX_K, 1)) + 
  theme(panel.grid.minor.x = element_blank(), panel.grid.minor.y = element_blank())
```
I will go with k=5 clusters

```{r}

pca_embeddings <- as.data.frame(pca$x[, 1:6])
# 1. Run Final K-Means with K = 5
set.seed(222) # Keeping your seed for reproducibility [cite: 1]
final_shooter_kmeans <- kmeans(pca_embeddings, centers = 4, nstart = 25)

# 2. Attach the cluster labels back to your original unscaled data
shot_data$Shooter_Cluster <- as.factor(final_shooter_kmeans$cluster)

# 3. Calculate the average stats for each cluster to see what they represent
cluster_profiles <- shot_data %>%
  group_by(Shooter_Cluster) %>%
  summarise(
    Count = n(),
    AVG_FGA =round(mean(FGA_per100), 1),
    Avg_USG = round(mean(USG.), 1),
    Avg_AST_per100 = round(mean(AST_per100), 1),
    Avg_Dist = round(mean(Dist), 1),
    Avg_3PAr = round(mean(X3PAr), 2),
    Avg_0_3ft = round(mean(pct_FGA_0_3), 3),
    AVG_Midrange = round(mean(pct_FGA_10_16), 3),
    Avg_Dunks = round(mean(pct_FGA_Dunks), 3),
    Avg_Ast_2P = round(mean(pct_Ast_2P), 3),
    Avg_Ast_3P = round(mean(pct_Ast_3P), 3),
    AVG_3PA_Corner = round(mean(pct_3PA_Corner), 3),
    # --- EFFICIENCY / SKILL ADDED ---
    Avg_FG_Rim = round(mean(fg_pct_0_3 , na.rm = TRUE), 3),
    Avg_FG_Short_Mid = round(mean(fg_pct_3_10 , na.rm = TRUE), 3),
    Avg_FG_Mid = round(mean(fg_pct_mid , na.rm = TRUE), 3),
    Avg_FG_Long_Mid = round(mean(fg_pct_long_mid   , na.rm = TRUE), 3),
    Avg_FG_3P = round(mean(fg_pct_3P , na.rm = TRUE), 3)
    
    
  )


# Print the profiles to the console
print(cluster_profiles)
write.csv(cluster_profiles, "cluster_profiles_15_16.csv", row.names = FALSE)


# 4. Export the mapping for Python / PyTorch
# We only need the Player name and their Cluster ID
shooter_export <- shot_data %>%
  select(Player, Shooter_Cluster)

write.csv(shooter_export, "shooter_archetypes_15_16.csv", row.names = FALSE)
```

Explanation of the clusters obtained:



```{r}

# 1. Get the MAP class labels from your PCA-trained GMM
km_labels <- final_shooter_kmeans$cluster

# 2. Bind those labels to your ORIGINAL scaled data (so we have basketball stats, not PCs)
km_profiles <- as.data.frame(scaled_data) %>%
  mutate(Cluster = paste("Cluster", km_labels))

  
km_centers <- km_profiles %>%
  group_by(Cluster) %>%
  summarise(across(everything(), mean))

# 2. Pivot to long format
library(tidyr)
km_centers_long <- km_centers %>% 
  pivot_longer(cols = -Cluster, names_to = "feature", values_to = "z_val")

# 3. The SOTA Z-Score Line Chart
library(ggplot2)
km_centers_long %>% 
  # Note the addition of 'group = Cluster' to allow geom_line to connect the dots
  ggplot(aes(x = reorder(feature, z_val), y = z_val, color = Cluster, group = Cluster)) + 
  geom_point(size = 3) + 
  geom_line(linewidth = 1) + 
  # Add the league average baseline
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) + 
  coord_flip() + # Flip to horizontal for readable feature names
  scale_color_brewer(palette = "Paired") + 
  facet_wrap(~ Cluster, ncol = 5) + # Put all 5 clusters side-by-side
  labs(title = "K-Means Archetype Profiles",
       subtitle = "Standard Deviations from NBA Average (Zero Line)",
       x = "Offensive Metric", 
       y = "Cluster Center (Z-score)") + 
  theme_minimal() + 
  theme(legend.position = "none", 
        strip.text = element_text(face = 'bold', size = 12),
        axis.text.y = element_text(size = 8), 
        panel.grid.minor = element_blank())
```





```{r}
library(dplyr)

# 1. Attach the K-Means cluster assignments to your dataset
# broom::augment would give us .cluster, but we can do it directly
kmodel_augment <- shot_data %>%
  mutate(.cluster = as.factor(final_shooter_kmeans$cluster))

# 2. Calculate the geometric distance from each player to their assigned centroid
# WARNING: We MUST use 'scaled_data' for the math, because that's what K-Means trained on!
assigned_centers <- final_shooter_kmeans$centers[final_shooter_kmeans$cluster, ]
distances <- sqrt(rowSums((pca_embeddings - assigned_centers)^2))

# Add the distance metric back to our dataframe
kmodel_augment$Distance_to_Centroid <- distances

# 3. Extract the Top 5 purest representatives (Smallest Distance)
top_5_kmeans_per_cluster <- kmodel_augment %>%
  dplyr::select(Player, .cluster, Distance_to_Centroid) %>%
  group_by(.cluster) %>%
  arrange(Distance_to_Centroid) %>%      # Ascending: Smallest distance = closest to the center
  slice_head(n = 5) %>%                  # Grab the top 5 closest players
  ungroup()                              # Always ungroup when done

# Print all 25 rows to the console
print(top_5_kmeans_per_cluster, n = 25)
```




```{r}
pc2 <- as.data.frame(pca$x[,1:2]) # extract first 3 PCs
pc2$Cluster <- as.factor(final_shooter_kmeans$cluster) # add player clusters 
cluster1_var <- round(pca_summary$importance[2,1], 4) * 100 # get variance explained by cluster 1
cluster2_var <- round(pca_summary$importance[2,2], 4) * 100 # get variance explained by cluster 2

# how different are the clusters when scaled down to two dimensions? 
pc2 %>% 
  ggplot(aes(x=PC1, y=PC2, color=Cluster, shape=Cluster)) + 
  geom_point(alpha=0.3) + 
  geom_rug() + # great way to visualize points on a single axis
  stat_ellipse(level=(2/3)) + # set ellipse value to one standard deviation
  scale_shape_manual(values=seq(0,15)) + 
  labs(x = paste0('PC1 (Accounts for ', cluster1_var, '% of Variance)'), # define cluster 1 % of variance
       y = paste0('PC2 (Accounts for ', cluster2_var, '% of Variance)'), # define cluster 2 % of variance
       title = 'Visualizing K-Means Cluster Differences in 2D') 
```
Gaussian mixture model for the archetype


```{r}
# run GMM on the scaled data from above
library(mclust)

set.seed(222)
mixture_model <-  Mclust(scaled_data, what = "BIC")
mix_model_summary <- summary(mixture_model) # obtain model summary
mix_model_summary
```

```{r}
plot(mixture_model, what = "BIC")
```




Lets now compare the Gaussian model with kmeans
```{r}
shot_data$KMeans_Cluster <- as.factor(final_shooter_kmeans$cluster)

# 2. Attach your new GMM clusters from the PCA
shot_data$GMM_Cluster <- as.factor(mixture_model$classification)

# 3. Create the cross-tabulation table to compare them
comparison_table <- table(KMeans = shot_data$KMeans_Cluster, 
                          GMM = shot_data$GMM_Cluster)

# Print the table
print(comparison_table)
```


```{r}


# 1. Attach the GMM cluster labels back to your original unscaled data
# $classification gives the "hard" assignment (the cluster with the highest probability)
# We use this primarily so we can group_by() and read the human-readable averages
shot_data$GMM_Cluster <- as.factor(mixture_model$classification)

# 2. Extract and attach the soft probabilities for XGBoost
gmm_probs <- as.data.frame(mixture_model$z)
# This dynamically names the columns Prob_Cluster_1, Prob_Cluster_2, etc., regardless of how many clusters you have
colnames(gmm_probs) <- paste0("Prob_Cluster_", 1:ncol(gmm_probs))

# Bind the probabilities to our main dataframe
shot_data <- cbind(shot_data, gmm_probs)

# 3. Calculate the average stats for each cluster using the NEW feature set
cluster_profiles <- shot_data %>%
  group_by(GMM_Cluster) %>%
  summarise(
    Count = n(),
    
    # --- OFFENSIVE GRAVITY & DIET ---
    Avg_USG = round(mean(USG., na.rm = TRUE), 1),
    Avg_X3PAr = round(mean(X3PAr, na.rm = TRUE), 3),
    
    # --- SHOT CREATION & SLASHING ---
    Avg_FTr = round(mean(FTr, na.rm = TRUE), 3),
    Avg_Ast_2P = round(mean(pct_Ast_2P, na.rm = TRUE), 3),
    Avg_Ast_3P = round(mean(pct_Ast_3P, na.rm = TRUE), 3),
    
    # --- EFFICIENCY / TOUCH ---
    Avg_FG_Rim = round(mean(fg_pct_0_3, na.rm = TRUE), 3),
    Avg_FG_Mid = round(mean(fg_pct_mid, na.rm = TRUE), 3), 
    Avg_FG_3P = round(mean(fg_pct_3P, na.rm = TRUE), 3)
    
    # Note: If you included FT_pct in your model, add it here:
    # Avg_FT_pct = round(mean(FT_pct, na.rm = TRUE), 3)
  )

# Print the profiles to the console so you can define your archetypes
print(cluster_profiles)

# Save the human-readable profiles for your thesis appendix
write.csv(cluster_profiles, "gmm_cluster_profiles_15_16.csv", row.names = FALSE)


# 4. Export the mapping for Python / XGBoost
# We select the Player, their hard cluster (for EDA plotting in Python), 
# and ALL the continuous probability columns for XGBoost to train on.
#shooter_export <- shot_data %>%
#  select(Player, GMM_Cluster, starts_with("Prob_Cluster_"))

#write.csv(shooter_export, "gmm_shooter_archetypes_15_16.csv", row.names = FALSE)
```




```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

# 1. Get the MAP class labels from your trained GMM
gmm_labels <- mixture_model$classification

# 2. Bind those labels to your ORIGINAL scaled data and assign Archetype names
gmm_profiles <- as.data.frame(scaled_data) %>%
  mutate(Cluster = factor(gmm_labels, 
                          levels = c(1, 2, 3, 4), 
                          labels = c("Primary Creators", 
                                     "Floor Spacers", 
                                     "Skilled Interior Hubs", 
                                     "Rim Runners")))

# 3. Calculate the true centers (means) for each feature by Cluster
mbc_centers <- gmm_profiles %>%
  group_by(Cluster) %>%
  summarise(across(everything(), mean))

# 4. Pivot for plotting
mbc_centers_long <- mbc_centers %>%
  pivot_longer(!Cluster, names_to = 'feature', values_to = 'z_val')

# 5. Plot the Z-Score Profiles
ggplot(mbc_centers_long, aes(x = feature, y = z_val, color = Cluster, group = Cluster)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  facet_wrap(~ Cluster, ncol = 4) + # Updated to 4 columns
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") + # 0 is the NBA average
  theme_minimal() +
  labs(title = "Offensive Archetype Profiles (Z-Scores)",
       x = "Offensive Feature",
       y = "Standard Deviations from NBA Average") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") # Hides redundant legend
```


```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Assuming mbc_centers_long is already loaded in your environment from the previous step
ggplot(mbc_centers_long, aes(x = feature, y = z_val, group = Cluster, color = Cluster, fill = Cluster)) +
  # Draw the filled polygon area
  geom_polygon(alpha = 0.3, linewidth = 1) +
  # Add points at the vertices
  geom_point(size = 2) +
  # Convert the standard x/y plot into a circular radar plot
  coord_polar() +
  # Create a separate radar chart for each archetype
  facet_wrap(~ Cluster, ncol = 2) +
  # Add the NBA average baseline (0 in Z-scores)
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  theme_minimal() +
  labs(
    title = "Offensive Archetype Signatures",
    subtitle = "Visualizing the feature footprint of each GMM cluster (Z-Scores)",
    x = NULL, 
    y = NULL
  ) +
  theme(
    axis.text.x = element_text(size = 8, face = "bold"),
    axis.text.y = element_blank(), # Hide the numerical Y axis for a cleaner look
    axis.ticks.y = element_blank(),
    panel.grid.major = element_line(color = "gray90"),
    legend.position = "none",
    strip.text = element_text(size = 12, face = "bold")
  )
```


```{r}
# Install if you don't have it: install.packages("ggridges")
library(ggridges)
library(ggplot2)
library(dplyr)
library(tidyr)

# 1. Select a few highly defining features from your gmm_profiles data
# (Assuming gmm_profiles has the original scaled stats and the named 'Cluster' column)
ridgeline_data <- gmm_profiles %>%
  select(Cluster, USG., X3PAr, FTr, pct_Ast_3P) %>% # Adjust these column names to exactly match your dataset
  pivot_longer(cols = -Cluster, names_to = "Feature", values_to = "Scaled_Value")

# 2. Clean up the feature names for the plot headers
ridgeline_data <- ridgeline_data %>%
  mutate(Feature = recode(Feature,
                          "USG." = "Usage Rate",
                          "X3PAr" = "3-Point Attempt Rate",
                          "FTr" = "Free Throw Rate",
                          "pct_Ast_3P" = "Assisted 3P%"))

# 3. Generate the Ridgeline Plot
ggplot(ridgeline_data, aes(x = Scaled_Value, y = Cluster, fill = Cluster)) +
  # scale = 1.2 allows the density curves to slightly overlap (the "ridgeline" effect)
  geom_density_ridges(alpha = 0.7, scale = 1.2, color = "white") +
  facet_wrap(~ Feature, scales = "free_x", ncol = 2) +
  theme_minimal() +
  labs(
    title = "Statistical Densities by Offensive Archetype",
    subtitle = "Visualizing how feature distributions overlap across the NBA landscape",
    x = "Standard Deviations from NBA Average (Z-Score)",
    y = NULL
  ) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 11, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    panel.grid.major.x = element_line(color = "gray80", linetype = "dashed"),
    panel.grid.minor = element_blank()
  )
```








```{r}
# changing the .class assignment in this chunk of code allows you to explore the players with the highest probability of belonging to that cluster
library(broom)
library(dplyr)
mix_model_preds <- as.data.frame(predict(mixture_model, scaled_data))
mix_model_augment <- augment(mixture_model, shot_data)
# Extract the Top 5 purest representatives for ALL 5 clusters instantly
top_5_per_cluster <- mix_model_augment %>%
  dplyr::select(Player, .class, .uncertainty) %>%
  group_by(.class) %>%
  arrange(.uncertainty) %>%      # Ascending: Lowest uncertainty = purest fit
  slice_head(n = 5) %>%          # Grab the top 5 from each group
  ungroup()                      # Always ungroup when done!

# Print all 25 rows to the console
print(top_5_per_cluster, n = 25)
```


```{r}
pc2 <- as.data.frame(pca$x[,1:2]) # extract first two PCs 
pc2$Cluster <- as.factor(mix_model_augment$.class) # add player clusters 


pc2 %>% 
  ggplot(aes(x=PC1, y=PC2, color=Cluster, shape=Cluster)) + 
  geom_point(alpha=0.5) +
  geom_rug() + theme_minimal() + 
  stat_ellipse(level=(2/3)) + # set ellipse value to one standard deviation
  scale_shape_manual(values=seq(0,7)) +
  labs(x = paste0('PC1 (Accounts for ', cluster1_var, '% of Variance)'),
       y = paste0('PC2 (Accounts for ', cluster2_var, '% of Variance)'),
       title = 'Visualizing Model Based Cluster Differences in 2D') 
```

```{r}
# Extract the probability matrix (N players x 5 clusters)
gmm_probs <- as.data.frame(mixture_model$z)

# Rename the columns so they make sense in Python
colnames(gmm_probs) <- c("Primary_Creator", "Spacer", "Mid-Interior", "Rim_Center")

# Bind the player names back to the probabilities
gmm_export <- cbind(Player = shot_data$Player, gmm_probs)

# Export for PyTorch/XGBoost
write.csv(gmm_export, "gmm_soft_labels_15_16.csv", row.names = FALSE)
```


```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# 1. Grab a sample of players (e.g., top 15 by Usage or just random notable names)
# For this example, we'll just take the first 15 rows, but you can filter for specific players
sample_players <- head(gmm_export, 35)

# 2. Pivot the data into long format for ggplot
gmm_long <- sample_players %>%
  pivot_longer(
    cols = c("Primary_Creator", "Spacer", "Mid-Interior", "Rim_Center"), 
    names_to = "Cluster", 
    values_to = "Probability"
  )

# 3. Clean up the cluster names for the legend
gmm_long$Cluster <- gsub("Prob_", "", gmm_long$Cluster)
gmm_long$Cluster <- gsub("_", " ", gmm_long$Cluster)

# 4. Plot the Stacked Bar Chart
ggplot(gmm_long, aes(x = reorder(Player, Probability), y = Probability, fill = Cluster)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() + # Flip horizontal to read names easily
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "GMM Soft Labels: Player Archetype DNA",
    subtitle = "Visualizing the Probability Matrix fed to PyTorch",
    x = "Player",
    y = "Cluster Probability"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(face = "bold")
  )
```



